---
title: "U6614 Final Project Proposal — Deliverable 2"
author: "Neil Chin (nkc2124) & Nicole Yahui Wu (yw3551)"
date: '`r Sys.Date()`'
output:
  pdf_document:
    toc: no
    toc_depth: '3'
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    df_print: tibble
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) 
library(tidycensus)
library(readxl)
library(lubridate)
```

# Research Question and Motivations

**Did 2014 Medicaid expansion reduce drug overdose deaths in Appalachia?**

Deaths due to drug overdose are a pressing problem facing United States policymakers, with reportedly more than 932,000 people dying due to overdose since 1999 ([CDC, 2022](https://www.cdc.gov/drugoverdose/deaths/index.html)). Rate of overdose deaths increased in nearly every US state from 2013-2017, with particularly severe incidence in the Appalachian region ([CDC, 2020](https://www.cdc.gov/drugoverdose/deaths/2013-2017-increase.html), [CDC, 2021](https://www.cdc.gov/drugoverdose/deaths/2014.html)). The increase in deaths is driven primarily by the ongoing US  opioid crisis, with the vast majority of deaths associated with opioid use ([CDC, 2022](https://www.cdc.gov/drugoverdose/deaths/index.html)).

In this context, we hope to assess whether expansion of the social safety net through government policy can be causally linked to reductions in drug overdose death rates. Specifically, we examine the estimated effects of the 2014 expansion of the US Medicaid program, which increased income-based eligibility for healthcare (including drug addiction treatment) in some Appalachian states, on drug overdose deaths. We hope that the findings from this study can assist policymakers in determining whether expanding access to healthcare in low-income areas is a viable policy mechanism for addressing the ongoing drug overdose crisis.

# Policy Background

The Affordable Care Act (ACA) was passed by the United States Congress and signed into law by President Barack Obama in 2010, drastically changing the policy landscape for health care in the United States. Among the major provisions in the ACA was expanded eligibility for Medicaid (i.e., "Medicaid Expansion"), which allowed states to raise the income-eligibility threshold to 138% of the federal poverty level ([KFF, 2022](https://www.kff.org/medicaid/issue-brief/status-of-state-medicaid-expansion-decisions-interactive-map/)). 

Among the 13 states whose boundaries overlap with the broad geographical definition of Appalachia, five states (Kentucky, Maryland, New York, Ohio, and West Virginia) passed legislation mandating the expansion of Medicaid as of January 1st, 2014 ([KFF, 2022](https://www.kff.org/medicaid/issue-brief/status-of-state-medicaid-expansion-decisions-interactive-map/)). Two additional states, Pennsylvania and Virginia, would later expand Medicaid, with the former in 2015 and the latter in 2019. Six states (Alabama, Georgia, Mississippi, North Carolina, South Carolina, and Tennessee) have not expanded Medicaid to-date. 

In our study, we identify this policy adoption discrepancy as a "treatment" (i.e., "differential exposure within entities over time") affecting drug overdose incidence in Appalachian communities. Our hypothesized causal mechanism is that expanded Medicaid eligibility allowed for greater access to low-cost health care among Appalachian counties in expansion states, therefore enabling people struggling with drug addiction to receive treatment when they otherwise would not have been able to receive care, reducing overall deaths from drug overdose in these areas.

# Data Description

Our data examines drug overdose deaths in Appalachian counties over the period 2012-2015. The subset of US counties defined as "Appalachian" is based on the jurisdiction of the Appalachian Regional Council ([ARC, 2021](https://www.arc.gov/appalachian-counties-served-by-arc/)). Accordingly, our units of observation for this study are counties, with the representative population being people living in the Appalachian region. We focus on counties as units of observation because the Appalachian region is defined at the county-level and to avoid potential selection bias and small sample size issues that would confound state-level analysis. 

Identification of state-level Medicaid Expansion is based on tracking done by the Kaiser Family Foundation ([KFF, 2022](https://www.kff.org/medicaid/issue-brief/status-of-state-medicaid-expansion-decisions-interactive-map/)). Due to the fact that Pennsylvania implemented Medicaid expansion a year after other expansion states, we exclude Pennsylvania counties from our analysis to ensure that there is a distinct time-based differential in treatment (i.e., pre- and post-2014 periods). Similarly, we consider Virginia to be a non-expansion state for the purposes of this study, because it only enacted Medicaid expansion after the end of our study period.

Data on drug overdose deaths comes from estimates modeled by the National Center for Health Statistics (NCHS), which are available at the county-level for the period 2003-2020. Unfortunately, county-level statistics on overdose deaths based on final counts of cause of death reporting only became available in 2020, outside of the time frame of this study.

County-level demographic covariates are taken from the US Census Bureau American Community Survey (ACS). Using these covariates, we hope to control for omitted variable bias stemming from county-level factors such as median income, median age, and sex and race compositions. In particular, we expect that median income would relate negatively to overdose deaths, as higher income communities would be better able to avoid drug addiction treatment, all else equal. We include controls for median age as drug overdose death incidence tends to vary by age group ([KFF, 2022](https://www.kff.org/other/state-indicator/opioid-overdose-deaths-by-age-group/?dataView=1&currentTimeframe=6&selectedDistributions=0-24--25-34--35-44--45-54&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)). Drug overdose deaths are also more common among individuals identified as male than female ([CDC, 2022](https://www.cdc.gov/nchs/hus/topics/drug-overdose-deaths.htm)).

# Preliminary Exploratory Analysis

Descriptive statistics for counties in expansion states and for counties in non-expansion states.

```{r}
year_county_panel <- readRDS("yearly_panel.rds")
```

[Note: Can use ggplot facet wrapping and histograms to show distribution between "treatment" and "control" groups for each year]

```{r}
year_county_panel %>% 
    ggplot(aes(x = estimated_overdose_death_rate, 
               color = as.factor(medicaid_expansion), 
               fill = as.factor(medicaid_expansion))) +
    geom_histogram() + 
    facet_wrap(~ Year, nrow = 4)
```

# Empirical Strategy

To evaluate the effect of Medicaid expansion on drug overdose deaths in Appalachian counties, we would like to estimate the following "differences-in-differences" specification:

$Overdose~Death~Rate_{it} =  Medicaid~Expansion_{it} + Medicaid~Expansion_{it}*Post~Expansion_{t} + Median~Income_{it} +Average~Age_{it} + Male~Share_{it} + Black~Share_{it} + White~Share_{it} + Hispanic~Share_{it} + Asian~Share_{it} + Post~Expansion_{t} + \Gamma_{i} + u_{it}$

[Explanation of why we use fixed effects]


Key limitations are the potential imbalance between pseudo "treatment" and "control" groups.

# Appendix

## Data Sourcing and Cleaning

### Appalachian Counties Data

The Appalachian Regional Commission defines 423 counties in 13 states (West Virginia, Alabama, Georgia, Kentucky, Maryland, Mississippi, New York, North Carolina, Ohio, Pennsylvania, South Carolina, Tennessee, and Virginia) as demarcating the Appalachian region. We adopt this geographical definition in our research approach.

More information available at: https://www.arc.gov/appalachian-counties-served-by-arc/.

```{r}
appalachian_counties <- read_excel('data/appalachian_counties_ARC_2021.xlsx')
```

### Medicaid Expansion

Of the 13 states, five states (Kentucky, Maryland, New York, Ohio, and West Virginia) passed legislation mandating the expansion of Medicaid as of January 1st, 2014. Two additional states, Pennsylvania and Virginia, would later expand Medicaid, with the former in 2015 and the latter in 2019. Six states (Alabama, Georgia, Mississippi, North Carolina, South Carolina, and Tennessee) have not expanded Medicaid to-date.

We add a dummy variable into our data to reflect this policy difference. Information on state-level Medicaid expansion available at: https://www.kff.org/medicaid/issue-brief/status-of-state-medicaid-expansion-decisions-interactive-map/.

```{r}
# Create dummy variable for medicaid expansion and a variable for date of expansion

expansion <- c("Kentucky", "Maryland", "New York", "Ohio", "West Virginia",
                     "Pennsylvania")
no_expansion <- c("Alabama", "Georgia", "Mississippi", "North Carolina", 
                  "South Carolina", "Tennessee", "Virginia")

appalachian_counties <- appalachian_counties %>% 
  mutate(medicaid_expansion = as.numeric(STATE %in% expansion),
         expansion_date = make_date(2014, 1, 1))

# Re-code dates for non-expansion and late expansion states

appalachian_counties$expansion_date[appalachian_counties$STATE %in% no_expansion] <-
  NA 
appalachian_counties$expansion_date[appalachian_counties$STATE == "Pennsylvania"] <-
  make_date(2015,1,1)
appalachian_counties$expansion_date[appalachian_counties$STATE == "Virginia"] <-
  make_date(2019,1,1)

```

### Overdose Deaths

Our data on drug overdose deaths comes from estimates modeled by the National Center for Health Statistics (NCHS), which are available at the county-level for the period 2003-2020. Source link: https://www.cdc.gov/nchs/data-visualization/drug-poisoning-mortality/.

```{r}
overdose_data <- read_csv("data/NCHS_-_Drug_Poisoning_Mortality_by_County__United_States.csv")

# Keep columns for FIPS codes, year, and modeled overdose deaths

overdose_data <- overdose_data %>% 
  select(FIPS, Year, `Model-based Death Rate`) %>%
  rename(estimated_overdose_death_rate = `Model-based Death Rate`)
```

### ACS Data

To control for time-variant, county-variant factors, we pull demographic data from the US Census Bureau's American Community Survey (ACS), for years 2012-2015. Data is downloaded from the US Census Bureau API using the package tidycensus.

```{r}
census_api_key("b79b301dc87bb0fd551147883a8141dca4e2823e")

# Pull 2012 ACS county-level data 

acs_2012 <- get_acs(geography = "county",
                   variables = c(population = "B01001_001",
                                 white_pop = "B01001H_001",
                                 hisp_pop = "B01001I_001",
                                 asian_pop = "B02001_005",
                                 black_pop = "B02001_003",
                                 male_pop = "B01001_002",
                                 med_income = "B19013_001",
                                 med_age = "B01002_001"),
                   year = 2012)

acs_2012 <- acs_2012 %>% 
  select(-moe, -NAME) %>% 
  spread(key = variable, value = estimate) %>% 
  mutate(Year = 2012,
         white_share = 100 * (white_pop/population),
         hisp_share = 100 * (hisp_pop/population),
         asian_share = 100 * (asian_pop/population),
         black_share = 100 * (black_pop/population),
         male_share = 100 * (male_pop/population))

# Pull 2013 ACS county-level data 

acs_2013 <- get_acs(geography = "county",
                   variables = c(population = "B01001_001",
                                 white_pop = "B01001H_001",
                                 hisp_pop = "B01001I_001",
                                 asian_pop = "B02001_005",
                                 black_pop = "B02001_003",
                                 male_pop = "B01001_002",
                                 med_income = "B19013_001",
                                 med_age = "B01002_001"),
                   year = 2013)

acs_2013 <- acs_2013 %>% 
  select(-moe, -NAME) %>% 
  spread(key = variable, value = estimate) %>% 
  mutate(Year = 2013,
         white_share = 100 * (white_pop/population),
         hisp_share = 100 * (hisp_pop/population),
         asian_share = 100 * (asian_pop/population),
         black_share = 100 * (black_pop/population),
         male_share = 100 * (male_pop/population))

# Pull 2014 ACS county-level data 

acs_2014 <- get_acs(geography = "county",
                   variables = c(population = "B01001_001",
                                 white_pop = "B01001H_001",
                                 hisp_pop = "B01001I_001",
                                 asian_pop = "B02001_005",
                                 black_pop = "B02001_003",
                                 male_pop = "B01001_002",
                                 med_income = "B19013_001",
                                 med_age = "B01002_001"),
                   year = 2014)

acs_2014 <- acs_2014 %>% 
  select(-moe, -NAME) %>% 
  spread(key = variable, value = estimate) %>% 
  mutate(Year = 2014,
         white_share = 100 * (white_pop/population),
         hisp_share = 100 * (hisp_pop/population),
         asian_share = 100 * (asian_pop/population),
         black_share = 100 * (black_pop/population),
         male_share = 100 * (male_pop/population))

# Pull 2015 ACS county-level data 

acs_2015 <- get_acs(geography = "county",
                   variables = c(population = "B01001_001",
                                 white_pop = "B01001H_001",
                                 hisp_pop = "B01001I_001",
                                 asian_pop = "B02001_005",
                                 black_pop = "B02001_003",
                                 male_pop = "B01001_002",
                                 med_income = "B19013_001",
                                 med_age = "B01002_001"),
                   year = 2015)

acs_2015 <- acs_2015 %>% 
  select(-moe, -NAME) %>% 
  spread(key = variable, value = estimate) %>% 
  mutate(Year = 2015,
         white_share = 100 * (white_pop/population),
         hisp_share = 100 * (hisp_pop/population),
         asian_share = 100 * (asian_pop/population),
         black_share = 100 * (black_pop/population),
         male_share = 100 * (male_pop/population))

# Append into single ACS data frame, rename GEOID as FIPS

acs_12_to_15 <- rbind(acs_2012, acs_2013, acs_2014, acs_2015)
acs_12_to_15 <- acs_12_to_15 %>% 
  mutate(GEOID = as.numeric(GEOID)) %>% 
  rename(FIPS = GEOID)
```

## Merge Data into Consolidated Data Set

We merge the Appalachian counties data, overdose data, and ACS data into a single data set.

```{r}

# Create four versions of the Appalachian counties data, one for each year in our study, then append into a panel

appalachian_counties_12 <- appalachian_counties %>% 
  mutate(Year = 2012)
appalachian_counties_13 <- appalachian_counties %>% 
  mutate(Year = 2013)
appalachian_counties_14 <- appalachian_counties %>% 
  mutate(Year = 2014)
appalachian_counties_15 <- appalachian_counties %>% 
  mutate(Year = 2015)

appalachian_counties_panel <- rbind(appalachian_counties_12,
                                    appalachian_counties_13,
                                    appalachian_counties_14,
                                    appalachian_counties_15)

# Merge panel with overdose data

appalachian_counties_overdose_panel <- appalachian_counties_panel %>% 
  mutate(FIPS = as.numeric(FIPS)) %>% 
  left_join(overdose_data, by = c("FIPS", "Year"))

# Merge panel with ACS data (save as an RDS file)

yearly_counties_panel_overdose_acs <- appalachian_counties_overdose_panel %>% 
  left_join(acs_12_to_15, by = c("FIPS", "Year"))

saveRDS(yearly_counties_panel_overdose_acs, file = "yearly_panel.rds")
```

